### 2017-01-18 abstract.tex
Formatting updated

Plasma irregularities in the ionosphere have been studied for decades using a variety of techniques.  A number of factors that contribute to the production and observation of plasma irregularities in the polar \(F\) region ionosphere are examined.  First, global factors related to both direct and indirect solar control are considered.  A band of ranges where echoes are statistically most likely to be observed is identified.  The location of this band varies diurnally and seasonally, but its motion relative to the radar can be explained by variations in the background plasma density effecting how far the beam must travel to be refracted to perpendicular to the magnetic field.  The high observed occurrence of nighttime backscatter implies an additional source of nighttime plasma density that is not accournted for by traditional models.  An example is presented that shows frequent polar patches may be the source of this density.  Backscatter occurrence peaks at the terminator.  Nighttime occurence is enhanced when a positive IMF By component is present, probably due to the convection pattern changing and creating a more favorable situation for gradient drift instability growth.

The gradient-drift instability is modeled within the field of view of a SuperDARN radar to identify when the growth rate is largest and examine asymmetry around a large-scale structure.  Dependencies on the relative directions of the gradient, plasma drift, and wavevector are all considered throughout the ionosphere.  In the \(F\) region, the strongest asymmetry is found when an elongated structure is oriented along the radar's boresight and moving perpendicular to its direction of elongation.  These results have implications for observations made with the SuperDARN network, but assume the directional dependencies predicted by linear GDI theory are applicable at decameter scales.

To test the accuracy of the predictions found from the model, irregularities surrounding a polar patch are analyzed in the context of linear GDI theory.  Backscatter power and occurrence from decameter scale irregularities are found using measurements from the SuperDARN radar at Rankin Inlet while background density gradients and electric fields are found from the north face of the Resolute Bay Incoherent Scatter Radar.  Particular emphasis is placed on directional dependence and wether small-scale irregularites show any signs of anisotropy.  While low powered echoes seem to follow the predictions of linear GDI theory, higher power echoes quickly exhibit much more complicated, nonlinear behavior.

>>>>

Plasma irregularities in the polar \(F\) region ionosphere are investigated primarily using ground based radars within the Super Dual Auroral Radar Network (SuperDARN) and the factors that contribute to the production and observation of these irregularities are considered.  First, solar control on irregularity production is evaluated.  A band of ranges where echoes are statistically most likely to be observed is identified.  The location of this band varies diurnally and seasonally, but its motion relative to the radar can be explained by background density with solar illumination effecting radar beam propogation.  The high observed occurrence of backscatter at night implies an additional source of nighttime plasma density that is not accounted for by traditional models.  An example is presented that shows frequent polar patches may be the source of this density.  Backscatter occurrence peaks at the terminator.  Nighttime occurence is enhanced when a positive IMF By component is present, probably due to changes in the convection pattern creating a more favorable situation for gradient-drift instability growth.

The gradient-drift instability is modeled within the field of view of a SuperDARN radar to identify when the growth rate is largest and examine asymmetry around large-scale structures, such as polar patches.  Dependencies on the relative directions of the gradient, plasma drift, and wavevector are all considered throughout the ionosphere.  In the \(F\) region, the strongest asymmetry is found when an elongated structure is oriented along the radar's boresight and moving perpendicular to its direction of elongation.  These results have implications for observations made with the SuperDARN network, but assume the directional dependencies predicted by linear GDI theory are applicable at decameter scales.

To test the accuracy of the predictions found from the model, irregularities surrounding a polar patch are analyzed in the context of linear gradient-drift instability theory.  Backscatter power and occurrence from decameter scale irregularities are found using measurements from the SuperDARN radar at Rankin Inlet while background density gradients and electric fields are found from the north face of the Resolute Bay Incoherent Scatter Radar.  Particular emphasis is placed on directional dependence and wether small-scale irregularities show any signs of anisotropy.  While low powered echoes seem to follow the predictions of linear theory, higher power echoes quickly exhibit much more complicated, nonlinear behavior.

### 2017-01-18 introduction.tex
Formatting updated

\begin{figure}
	\includegraphics[width=\textwidth,angle=0]{ste.pdf}
	\caption{Diagram of the solar-terrestrial environment.  The solar wind contains both a steady stream of charged particles and the IMF, dark blue.  The red region represents the magnetosheath.  The manetopause is shown by the black dashed line.  Closed magnetic field lines from the earth are shown in grey while open field lines are shown in light blue.}
	\label{fig:ste}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=\textwidth]{densprofile.pdf}
	\caption{Altitude profile of electron density throughout the ionosphere found from the IRI model.  Typical ranges of the \(D\), \(E\), and \(F\) regions are identified.}
	\label{fig:densprofile}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=0.8\textwidth]{irregularity.pdf}
	\caption{Schematic of different scale size plasma density irregularities that occur in the polar cap.  Polar patches, orange, move with the background ionospheric convection, indicated by green contours.  Field-aligned irregularities, FAIs, are density perturbations that form along the magnetic field lines, pink lines.}
	\label{fig:polarirreg}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=\textwidth,angle=180]{SuperDARNmap.pdf}
	\caption{FoVs of all operational SuperDARN radars in both the northern (left) and southern (right) hemispheres.  Polar latitude radars are shown with the red outlines.  The radars at Rankin Inlet (northern hemisphere) and McMurdo Station (southern hemisphere) are shaded in red because these two instruments are particularly important in the following studies.  Lines of constant magnetic latitude are shown in yellow at \(\Lambda = 40\deg, 60\deg, 80\deg\) in the northern hemisphere and at \(\Lambda = -40\deg,-60\deg,-80\deg\) in the southern hemisphere.}
	\label{fig:superdarnmap}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=\textwidth]{ISRmap.pdf}
	\caption{FoVs of all ISRs in the northern polar region.  The individual radars shown are identified by their commonly used acronyms: Poker Flat (PFISR), Resolute Bay North (RISR-N), Resolute Bay Canada (RISR-C), Sondrostrom (SFJ).  RISR-N, which is primarially used throughout this thesis, is shown in orange.  The FoV of the RKN SuperDARN radar is shown by the red outline.  Lines of constant MLAT are shown at \(\Lambda=80\deg,60\deg\) in yellow.}
	\label{fig:isrmap}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=0.75\textwidth,angle=270]{magnetosphere.pdf}
	\caption{Diagram of the interactions between the IMF, magnetosphere, and polar cap ionosphere.  Closed field lines in the Earth's magnetosphere are shown in grey while open field lines are light blue.  The polar cap open-closed boundary is in green.  Magnetic field lines that are part of the IMF are shown in dark blue.  Magnetic reconnection occurs along the X-line (red) on the dayside of the magnetopause (dashed line).  The X-line can be mapped back along magnetic field lines (orange) to the ionosphere, where their projection is the merging gap (pink dotted line).  During reconnection events, the open-closed boundary expands equatorwards as more open field lines are formed and there is plasma flow across the merging gap.}
	\label{fig:magnetosphere}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=0.75\textwidth,angle=270]{patch_formation.pdf}
	\caption{Polar patch formation through transient magnetopause reconnection.  The open-closed boundary is the green line.  Black contours represent plasma flow.  Dense daytime plasma is shown by the orange regions.  The burst of reconnection is occurring in Figures \ref{fig:patch_formation}a--\ref{fig:patch_formation}c but stops in Figures \ref{fig:patch_formation}d--\ref{fig:patch_formation}f.}
	\label{fig:patch_formation}
\end{figure}

@@

\begin{table}
\begin{tabular}{l l}
Limiting Case & Growth Rate, \(\gamma\) \\
\hline
GDI/FBI in \(E\) Region & \(\gamma = \frac{1}{1+\hat{\psi}}\left[\frac{\hat{\psi}}{\nu_i}\left(\left(\frac{\vec{V}_d\cdot \vec{k}}{1+\hat{\psi}}\right)^2-C_s^2 k_\perp^2\right)+\frac{br_i\vec{V}_d\cdot \vec{k}}{1+\hat{\psi}}\right]\) \\
GDI/CCI in \(F\) Region & \(\gamma = -\frac{b}{\psi+y^2}\left[\frac{\psi\vec{k}\cdot\vec{E}_{0\perp}+E_{0\parallel}k_\parallel}{B}\right]+C\left[r_ek_\perp^2-\frac{k_\parallel^2}{r_i}\left(1+\frac{r_i^2}{\psi+y^2}\right)\right]\) \\
Cold Plasma in \(F\) Region & \(\gamma = \frac{b}{1+\psi}\left(R\vec{V}_E-\frac{\vec{E}_{0\perp}}{B}\right)\cdot\vec{k}\) \\
\end{tabular}
\caption{Irregularity growth rates in the ionosphere for three different limiting cases \citep{Makarevich2016a}.}
\label{tab:gdi_exps}
\end{table}

>>>>

\begin{figure}
	\includegraphics[width=\textwidth,angle=0]{ste.pdf}
	\caption[Solar-Terrestrial Environment]{Diagram of the solar-terrestrial environment.  The solar wind contains both a steady stream of charged particles and the IMF, dark blue.  The red region represents the magnetosheath.  The manetopause is shown by the black dashed line.  Closed magnetic field lines from the earth are shown in grey while open field lines are shown in light blue.}
	\label{fig:ste}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=\textwidth]{densprofile.pdf}
	\caption[Ionosphere density profile]{Altitude profile of electron density throughout the ionosphere found from the IRI model.  Typical ranges of the \(D\), \(E\), and \(F\) regions are identified.}
	\label{fig:densprofile}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=0.8\textwidth]{irregularity.pdf}
	\caption[Plasma irregularities in the polar ionosphere]{Schematic of different scale size plasma density irregularities that occur in the polar cap.  Polar patches, orange, move with the background ionospheric convection, indicated by green contours.  Smaller-scale structures often move develop along the edges of polar patches where strong density gradients exist.  Field-aligned irregularities, FAIs, are density perturbations that form along the magnetic field lines, pink lines.}
	\label{fig:polarirreg}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=\textwidth,angle=180]{SuperDARNmap.pdf}
	\caption[SuperDARN map]{FoVs of all operational SuperDARN radars in both the northern (left) and southern (right) hemispheres.  Polar latitude radars are shown with the red outlines.  The radars at Rankin Inlet (northern hemisphere) and McMurdo Station (southern hemisphere) are shaded in red because these two instruments are particularly important in the following studies.  Lines of constant magnetic latitude are shown in yellow at \(\Lambda = 40\deg, 60\deg, 80\deg\) in the northern hemisphere and at \(\Lambda = -40\deg,-60\deg,-80\deg\) in the southern hemisphere.}
	\label{fig:superdarnmap}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=\textwidth]{ISRmap.pdf}
	\caption[ISR map]{FoVs of all ISRs in the northern polar region.  The individual radars shown are identified by their commonly used acronyms: Poker Flat (PFISR), Resolute Bay North (RISR-N), Resolute Bay Canada (RISR-C), Sondrostrom (SFJ).  RISR-N, which is primarially used throughout this thesis, is shown in orange.  The FoV of the RKN SuperDARN radar is shown by the red outline.  Lines of constant MLAT are shown at \(\Lambda=80\deg,60\deg\) in yellow.}
	\label{fig:isrmap}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=0.75\textwidth,angle=270]{magnetosphere.pdf}
	\caption[Magnetic reconnection footprint in the ionosphere]{Diagram of the interactions between the IMF, magnetosphere, and polar cap ionosphere.  Closed field lines in the Earth's magnetosphere are shown in grey while open field lines are light blue.  The polar cap open-closed boundary is in green.  Magnetic field lines that are part of the IMF are shown in dark blue.  Magnetic reconnection occurs along the X-line (red) on the dayside of the magnetopause (dashed line).  The X-line can be mapped back along magnetic field lines (orange) to the ionosphere, where their projection is the merging gap (pink dotted line).  During reconnection events, the open-closed boundary expands equatorwards as more open field lines are formed and there is plasma flow across the merging gap. Figure adapted from \citet{Cowley1991}.}
	\label{fig:magnetosphere}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=0.75\textwidth,angle=270]{patch_formation.pdf}
	\caption[Polar patch formation]{Polar patch formation through transient magnetopause reconnection.  The open-closed boundary is the green line.  Black contours represent plasma flow.  Dense daytime plasma is shown by the orange regions.  The burst of reconnection is occurring in Figures \ref{fig:patch_formation}a--\ref{fig:patch_formation}c but stops in Figures \ref{fig:patch_formation}d--\ref{fig:patch_formation}f.  Figure adapted from \citet{Cowley1991}.}
	\label{fig:patch_formation}
\end{figure}

@@

\begin{table}
\begin{tabular}{l l}
Limiting Case & Growth Rate, \(\gamma\) \\
\hline
GDI/FBI in \(E\) Region & \(\gamma = \frac{1}{1+\hat{\psi}}\left[\frac{\hat{\psi}}{\nu_i}\left(\left(\frac{\vec{V}_d\cdot \vec{k}}{1+\hat{\psi}}\right)^2-C_s^2 k_\perp^2\right)+\frac{br_i\vec{V}_d\cdot \vec{k}}{1+\hat{\psi}}\right]\) \\
GDI/CCI in \(F\) Region & \(\gamma = -\frac{b}{\psi+y^2}\left[\frac{\psi\vec{k}\cdot\vec{E}_{0\perp}+E_{0\parallel}k_\parallel}{B}\right]+C\left[r_ek_\perp^2-\frac{k_\parallel^2}{r_i}\left(1+\frac{r_i^2}{\psi+y^2}\right)\right]\) \\
Cold Plasma in \(F\) Region & \(\gamma = \frac{b}{1+\psi}\left(R\vec{V}_E-\frac{\vec{E}_{0\perp}}{B}\right)\cdot\vec{k}\) \\
\end{tabular}
\caption[Irregularity growth rates]{Irregularity growth rates in the ionosphere for three different limiting cases \citep{Makarevich2016a}.}
\label{tab:gdi_exps}
\end{table}
